Causal Analysis Paper

### 2022-06-07 PaperAnalysis.Rmd
Update text - Identification

## Introduction

@@

## Description of the original paper

@@

##  Methodology

### Theorethical estimands

@@

$$
\mathbb{E}[MMR^{0} \mid L = 1] = \mathbb{E}[MMR^{0} \mid L = 0]
$$
* The treatment effect will not be different in the treated and the untreated group, that is, countries that have inflexible abortion laws would do *as good* as countries with flexible abortion laws, if they made the laws more flexible:

@@

### Identification

@@

The current paper proposes that flexibility of abortion laws causes maternal mortality, and that GDP and female primary education completion are confounders of this relationship -- that is, they cause both whether or not a country will have flexible abortion laws and how big MMR will be. They also propose that the causal mechanism by which this relationship comes about is that the women in countries where restrictive laws are in place are less able (or unable) to access reproductive health care and safe abortion services, which increases the proportion of unsafe abortions, and in turn, maternal deaths. The model can be seen in the figure below:

@@

![Figure 1: Causal model as recreated from @Latt2019 ](Figures/dag1.png)

@@

####Unmeasured counfounding

As this is an observational study, where the treatment is not assigned in a random way but rather observed, there are several assumptions that need to be made to be able to estimate causal effects.

First, effects cannot be estimated if there is unmeasured counfounding. This means that all common causes of treatment assignment and the outcome must be included in the analysis.

####Positivity assumption

The second assumption is that of positivity, meaning that all units of analysis should have a nonzero probability of receiving either treatment.

####SUTVA

Finally, the third requirement is the stable unit treatment value assumption (SUTVA), meaning that there are no unmeasured versions of treatment and that the treatment status of any other unit does not affect other unit's outcomes.

### Empirical estimation

####Heterogeinity of effects

####G-separation

### Unmeasured confounding

Possible confounders that where not considered (directly) in the original paper

Health expenditure: health expenditure likely determines the access to safe abortions and women’s trust in medical institutions. The original paper found that GDP and health expenditure are so closely related that they used GDP as a proxy. We will keep this assumption.

Religious views in society and level of secularism: The power of religious institutions and they views they hold in regard to the acceptability of abortion is likely to influence the level of maternal mortality and the flexibility of abortion law. If for instance the church has a lot of power in society and is very traditional this likely will influence the flexibility of the law but also the individual’s choice whether to get a clandestine abortion outside of the health system or she will even have the possibility to acquire a safe abortion in a health institution. However, this confounder would be extremely difficult to measure as there is no systematic recording of religious power and the views different institutions hold for all the studied countries.

Party in power politically: The political convictions and policies of political parties in power most likely will influence the abortion law. However, it is unlikely that it will directly cause mortality. Therefore, it is not a confounder in this model.

@@

```{r}
dag  <- dagify(MMrates ~ laws,
               MMrates ~ GDP,
               MMrates ~ literacy,
               laws ~ GDP,
               laws~literacy
               )
ggdag(dag) +  theme_dag_blank()
```
However, it is also possible that female literacy has no effect on abortion law and actually only has an effect on the outcome, MMR. Then instead of causal effect it would be an associative risk factor.(Hernandez p.3). In this case the only connection would be if we conditioned on MMR we would create a spurious relation however the effect of literacy and law on each other would be zero. MMR is a common effect of female literacy and flexibility score (abortion laws).
```{r}
dag  <- dagify(MMrates ~ laws,
               MMrates ~ GDP,
               MMrates ~ literacy,
               laws ~ GDP
               )
ggdag(dag) +  theme_dag_blank()
```
However, it is difficult to disentangle the relationship as lower female literacy will also lower access to safe abortions regardless of the flexibility of the law as women are likely less informed about their health care options which in turn increases female mortality.

@@

Hernandez is outlining in their paper why data is missing, to restrict the analysis on countries where data are complete would bias it. Countries with missing data on female literacy are still included. (p. 5, Hernandez at all.)Female primary completion is only available for 50 percent of countries and female literacy only for 10 percent. This missingness of data introduces selection bias to this specific analysis the auhtors did in the original paper. As the likely selection bias in this study cannot be avoided due to the limitations on data availability it needs to be corrected if possible, in the analysis (Hernan p. 7) this could be achieved through generalization of inverse probability weighting. For this however we would need to categorize developing countries in one group where some have recorded female literacy which then gets multiplied by the number of countries not considered due to missing data. Inverse probability weighting would work as the score for female literacy is not zero for the countries where it is not recorded.  However, a lot of assumptions would need to be made to apply this method. The assumption that the filler-country would be close enough to the unobserved countries is by nature untestable and therefor it is a strong assumption.
=======
>>>>>>> 2390f3bced1f46e03a643063dda253b0010c6416
## Conclusion 
The policy conclusion the original paper draws is justified and we could rely on this to give advice to our clients. However, we could expand upon the methodology to get a more precise result and consider possible models as the limitation of models is not sufficiently justified in their paper.

>>>>

# Introduction

@@

# Description of the original paper

@@

#  Methodology

## Theorethical estimands

@@

$$
\mathbb{E}[MMR^{0} \mid L = 1] = \mathbb{E}[MMR^{0} \mid L = 0]
$$

* The treatment effect will not be different in the treated and the untreated group, that is, countries that have inflexible abortion laws would do *as good* as countries with flexible abortion laws, if they made the laws more flexible:

@@

## Identification

@@

The current paper proposes that flexibility of abortion laws causes maternal mortality, and that GDP and female primary education completion are confounders of this relationship -- that is, they cause both whether or not a country will have flexible abortion laws and how big MMR will be. The authors also propose that the causal mechanism by which this relationship comes about is that the women in countries where restrictive laws are in place are less able (or unable) to access reproductive health care and safe abortion services, which increases the proportion of unsafe abortions, and in turn, maternal deaths. The model can be seen in the figure below:

@@

![Figure 1.: Causal model as recreated from @Latt2019 ](Figures/dag1.png)

@@

### Unmeasured counfounding

What makes the estimation of ATE plausible is “uncoupling” or *d-separation* of assignment of units of analysis (in this case, countries) into treatment and control conditions. In randomized trials, this is achieved by randomly assigning treatment to unit analysis (Figure 2a). Randomization:

* Identifies the effect of Abortion law on MMR;
* Generated approximate balance in counfounders, even if a counfounder is unobserved;

which means that the causal effect of Law on MMR can be estimated. Conditioning on counfounders is still (statictically) desirable, as it will increase the precision of the effect estimate.

```{r, echo=FALSE,out.width="49%",out.height="49%",fig.show='hold', fig.align='center', fig.cap="Figure 2. a) Randomisation removes inbound connections from confounders to treatment, b) Without controling for all counfounders, the causal effect can't be estimated"}

knitr::include_graphics(c("Figures/dag2a.png","Figures/dag2b.png"))
```

However, in observational studies, the causal effect is identified only when we control for all counfounders. If there is unobserved counfounding (Figure2b), the causal effect can't be identified as: 
$$
MMR^{(1)}, MMR^{(0)} \indep Law \mid Female \ primary \ education, GDP, Unmeasured \ counfounders
$$ 
Given that unmeasured confounding is a crucial issue in effect identification, we have considered possible omitted variables and their role in the original model:

* Health expenditure: health expenditure likely determines the access to safe abortions and women’s trust in medical institutions. The original paper found that GDP and health expenditure are so closely related that they used GDP as a proxy. We will keep this assumption.

* Religious views in society and level of secularism: The power of religious institutions and they views they hold in regard to the acceptability of abortion is likely to influence the level of maternal mortality and the flexibility of abortion law. If for instance the church has a lot of power in society and is very traditional this likely will influence the flexibility of the law but also the individual’s choice whether to get a clandestine abortion outside of the health system or she will even have the possibility to acquire a safe abortion in a health institution. However, this confounder would be extremely difficult to measure as there is no systematic recording of religious power and the views different institutions hold for all the studied countries.

* Party in power politically: The political convictions and policies of political parties in power most likely will influence the abortion law. However, it is unlikely that it will directly cause mortality. Therefore, it is not a confounder in this model.

* State of human rights and gender equality: State of human rights and gender equality in a country could influence the abortion law flexibility and female primary education completion. However, we think it is unlikely that they will impact MMR; therefore, it would not be necessary to control for them, as long as female primary education is controlled for.

@@

However, it is also possible that female literacy has no effect on abortion law and actually only has an effect on the outcome, MMR. Then instead of causal effect it would be an associative risk factor.(Hernandez p.3). In this case the only connection would be if we conditioned on MMR we would create a spurious relation however the effect of literacy and law on each other would be zero. MMR is a common effect of female literacy and flexibility score (abortion laws).

```{r}
dag  <- dagify(MMrates ~ laws,
               MMrates ~ GDP,
               MMrates ~ literacy,
               laws ~ GDP
               )
ggdag(dag) +  theme_dag_blank()
```

However, it is difficult to disentangle the relationship as lower female literacy will also lower access to safe abortions regardless of the flexibility of the law as women are likely less informed about their health care options which in turn increases female mortality.

@@

Hernandez is outlining in their paper why data is missing, to restrict the analysis on countries where data are complete would bias it. Countries with missing data on female literacy are still included. (p. 5, Hernandez at all.)Female primary completion is only available for 50 percent of countries and female literacy only for 10 percent. This missingness of data introduces selection bias to this specific analysis the auhtors did in the original paper. As the likely selection bias in this study cannot be avoided due to the limitations on data availability it needs to be corrected if possible, in the analysis (Hernan p. 7) this could be achieved through generalization of inverse probability weighting. For this however we would need to categorize developing countries in one group where some have recorded female literacy which then gets multiplied by the number of countries not considered due to missing data. Inverse probability weighting would work as the score for female literacy is not zero for the countries where it is not recorded.  However, a lot of assumptions would need to be made to apply this method. The assumption that the filler-country would be close enough to the unobserved countries is by nature untestable and therefor it is a strong assumption.

### Positivity assumption

?
The second assumption is that of positivity, meaning that all units of analysis should have a nonzero probability of receiving either treatment.

### SUTVA

?
Finally, the third requirement is the stable unit treatment value assumption (SUTVA), meaning that there are no unmeasured versions of treatment and that the treatment status of any other unit does not affect other unit's outcomes.

## Empirical estimation

### Heterogeinity of effects

### G-separation

# Analysis

* Dataset description
* Regression ATE
* Heterogeinity of effects
* Unmeasured confounding
* G-separation ATE, ATU, ATT

# Conclusion 
The policy conclusion the original paper draws is justified and we could rely on this to give advice to our clients. However, we could expand upon the methodology to get a more precise result and consider possible models as the limitation of models is not sufficiently justified in their paper.
